---
title: Water
subtitle: Long-Term Trends in Groundwater Levels in B.C.
output:
  pdf_document:
    template: D:/templates/print_ver_template.tex
header-includes:
  - \usepackage{xcolor,colortbl}
  - \usepackage{longtable, booktabs}
---

```{r set-options, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library("dplyr")
library("ggplot2")
library("grid")
library("RColorBrewer")
library("scales")
library("ggmap")
library("bcgroundwater")
library("envreportutils")

base_theme <- theme_get()

load('../tempdata/analysisData.RData')
load('../tempdata/mapData.RData')

attr_wells <- attr.out[attr.out$category != "N/A",]

attr_wells$state <- factor(attr_wells$state
                         , levels=c("Increasing", "Stable"
                                    , "Moderate rate of decline"
                                    , "Large rate of decline")
                         , ordered=TRUE)

attr_wells$category <- factor(attr_wells$category, 
                            levels=c("Stable or Increasing", 
                                     "Moderate rate of decline"
                                     , "Large rate of decline")
                            , ordered=TRUE)

state.summary <- table(attr_wells$state)
state.summary <- cbind(n=state.summary, prop=prop.table(state.summary))

theme_set(theme_classic() + 
            theme(text=element_text(colour="#666666")
                  , axis.line=element_blank()
                  , axis.ticks=element_blank()
                  , panel.grid.major=element_line(colour="grey85", size=0.5
                                                  , linetype=1)
                  , panel.grid.minor=element_line(colour="grey90", size=0.5
                                                  , linetype=1)
                  , panel.grid.major.x=element_blank()
                  , panel.margin=unit(0.6, "lines")
                  , plot.title=element_text(vjust=2)
                  , axis.title=element_text(vjust=0.1)
                  , legend.position="bottom", legend.title=element_blank()
                  , axis.text.x = element_blank()
                  , strip.background=element_blank()))

label.colour <- "#3d3d3d"
colour.scale <- brewer.pal(3,"Blues")

updateDate <- format(Sys.Date(), "%B %Y")
```

Introduction
============

Up to 1 million British Columbians are estimated to consume groundwater,
and hundreds of groundwater aquifers provide water for industries,
municipalities, farms, and rural homeowners in B.C. British Columbia
operates a provincial observation well network of over 180 wells, which
was established in 1961 to monitor groundwater availability in areas of
high human use.

-   Observation wells are not used for domestic or commercial use, but
    instead provide information on groundwater levels over time.

-   Monitoring groundwater levels allows us to know how much groundwater
    is available given human use patterns, aquifer characteristics,
    weather and climate patterns.

-   This indicator presents a statistical analysis of long-term trends
    in groundwater levels recorded at `r nrow(attr_wells)` observation wells 
    that have been monitored for ten years or more and were active as of 2004.

-   `r round(sum(state.summary[c("Stable","Increasing"),"prop"]*100))`% of 
    observation wells examined have water levels that are stable
    or increasing (with `r state.summary["Increasing","n"]` wells showing 
    increasing trends); `r round(state.summary["Large rate of decline","prop"]*100)`% 
    of observation wells show a large rate of decline in water levels (more
    than 10 cm per year), with a further 
    `r round(state.summary["Moderate rate of decline","prop"]*100)`% of wells 
    showing a moderate rate of decline in water levels (between 3 and 10 cm per year).
    
Provincial summaries
====================

The charts below summarise the proportions of observation wells with
groundwater levels in three long-term trend categories: large rate of
decline (more than 10 cm/year), moderate rate of decline (3 to 10
cm/year), and stable or increasing. These summaries are also presented
by region and aquifer type.

```{r pie, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, fig.width=5, fig.height=6, fig.align='center'}
sum.data <- attr_wells %>%
  group_by(category) %>%
  summarise(Freq = n()) %>%
  mutate(per=round(Freq/sum(Freq)*100)
         , pos=cumsum(Freq)-Freq/2)

ggplot(attr_wells, aes(x=factor(1), fill=category)) + 
  geom_bar(width=1) + coord_polar(theta="y") + 
  scale_fill_manual(values=colour.scale) + 
  theme(line=element_blank(), axis.text=element_blank()
        , axis.title=element_blank(), plot.title=element_text(vjust=0)
        , legend.position=c(0.5,0), legend.direction="vertical"
        , legend.title=element_blank(), plot.margin=unit(c(rep(0,4)), "cm")) + 
  geom_text(data=sum.data
            , aes(x=1.2, y=pos, label=paste0(per,"%"))
            , colour=label.colour) + 
  labs(title=paste("Percentage of observation wells in three categories", 
                "of long-term trends in groundwater levels", sep="\n"))
```


```{r summaries, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
nLabeller <- function(n, singular, sep=" ") {
  suffix <- ifelse(n == 1, singular, paste0(singular,"s"))
  label <- paste(n, suffix, sep=sep)
  label
}

sum.data.reg <- attr_wells %>%
  group_by(REGION_NM, category) %>%
  summarise(Freq = n()) %>%
  mutate(prop=Freq/sum(Freq), 
         region_lab = paste0(gsub("\\s","\\\n", 
                                  gsub("\\s/\\s*", "/\\\n", REGION_NM)), 
                             "\n(", nLabeller(sum(Freq), "well"), ")"))

## Plot with percentage on y and sample size labels
reg.plot <- ggplot(sum.data.reg, aes(x=category, y=prop, fill=category)) + 
  geom_bar(stat='identity') + facet_grid(~ region_lab) + 
  labs(title="Variation in long-term trends in groundwater levels by region", 
       x=element_blank(), y="Percent of wells") + 
  scale_fill_manual(values=colour.scale) +
  scale_y_continuous(labels=percent, limits=c(0,1)) # + 

# Summarize by aquifer type
sum.data.aq <- attr_wells[attr_wells$Aquifer_Type != "Unknown",] %>%
  group_by(Aquifer_Type, category) %>%
  summarise(Freq = n()) %>%
  mutate(prop=Freq/sum(Freq), 
         aq_lab = paste0(Aquifer_Type, "\n(", 
                         nLabeller(sum(Freq), "well"), ")"))

## Plot with percentage on y and sample size labels
aq.plot <- ggplot(sum.data.aq, aes(x=category, y=prop, fill=category)) + 
  geom_bar(stat='identity') + facet_grid(~ aq_lab) + 
  labs(title="Variation in long-term trends in groundwater levels by aquifer type", 
       x=element_blank(), y="Percent of wells") + 
  scale_fill_manual(values=colour.scale) +
  scale_y_continuous(labels=percent, limits=c(0,1))

plot(reg.plot)
plot(aq.plot)
```

Individual well statistics
==========================

Groundwater levels are sensitive to precipitation, aquifer storage
capacity, recharge rate, and human withdrawal. Groundwater level trends
presented here indicate long-term changes in water level, but have not
been corrected for changes in precipitation patterns or other factors.
Thus, any significant trends are not necessarily directly attributable
to human use. However, information on long-term trends can be useful for
prompting further research and informing decision-making.

Locations of `r nrow(attr_wells)` groundwater observation wells that were included in the analysis
---------------------------------------------------------------------------------

```{r overview_map, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, results='asis', fig.height=6}
ggmap(ggMapBC, extent="device") + 
  coord_map(xlim=c(-139, -114), ylim=c(47.8,60)) + 
  geom_point(data=attr_wells, aes(x=Long, y=Lat, fill=category), 
             shape=21, size=2.5, colour=colour.scale[3]) + 
  scale_fill_manual(values=colour.scale) + 
  theme(legend.position="bottom", legend.title=element_blank())
```
\newpage

\definecolor{stable}{HTML}{`r substr(colour.scale[1], 2, 7)`}
\definecolor{moderate}{HTML}{`r substr(colour.scale[2], 2, 7)`}
\definecolor{large}{HTML}{`r substr(colour.scale[3], 2, 7)`}

\small
```{r table, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, results='asis'}
require(xtable)
require(lubridate)

out.table <- attr_wells[,c("REGION_NM", "Well_Num", "start_date", "last_date",  
                         "trend_line_slope", "sig", "state")]

out.table$date_range <- paste(year(out.table$start_date), "-", 
                              year(out.table$last_date))

out.table <- arrange(out.table, REGION_NM, as.numeric(Well_Num))

out.table$trend_line_slope <- -out.table$trend_line_slope

out.table <- out.table %>%
  select(REGION_NM, Well_Num, date_range, trend_line_slope, sig, state)

names(out.table) <- c("Region", "Well", "Date Range", 
                      "Slope (m/yr)", "Sig", "State")

out.table$Sig <- ifelse(out.table$Sig < 0.05, "p < 0.05", 
                        round(out.table$Sig,3))

x.table <- xtable(out.table, digits=2)

# Set the slope column as fixed width
align(x.table) <- c('l', rep('l', 3), 'p{0.5in}', rep('l', 2))

# Define colours of rows based on state
HL_col <- ifelse(out.table$State == "Large rate of decline", "\\rowcolor{large}", 
                 ifelse(out.table$State == "Moderate rate of decline", 
                        "\\rowcolor{moderate}", "\\rowcolor{stable}"))
HL_rows <- seq(0, length(HL_col)-1, by=1)
HL_col[1] <- paste0("\\hline ", HL_col[1]) # Hack to get a hline below headings

print(x.table, tabular.environment='longtable', floating=FALSE,
      hline.after=c(-1,nrow(x.table)), 
      add.to.row=list(pos=list(as.list(HL_rows))[[1]], 
                      command=HL_col), 
      booktabs=TRUE, 
      include.rownames=FALSE, sanitize.colnames.function=NULL, 
      sanitize.text.function=NULL, comment = FALSE)
```

References and Other Useful Links
=================================

-   Provincial Observation Well Network homepage:
    <http://www.env.gov.bc.ca/wsd/data_searches/obswell/index.html>

-   WELLS - Ground Water Wells and Aquifer Database:
    <http://www.env.gov.bc.ca/wsd/data_searches/wells/>

-   B.C. Water Information and Tools:
    <http://www2.gov.bc.ca/gov/topic.page?id=F281F036D5394460B4C3CC01F47833A7>

-   Wei, M., D. Allen, A. Kohut, S. Grasby, K. Ronneseth, and B. Turner.
    2009. Understanding the Types of Aquifers in the Canadian Cordillera
    Hydrogeologic Region to Better Manage and Protect Groundwater.
    Streamline Water Management Bulletin. 13 (1): 10-18.
    <http://forrex.org/sites/default/files/publications/articles/Streamline_Vol13_No1_Art3.pdf>

Data
====

\*By accessing these datasets, you agree to the license associated with
each file, as indicated in parentheses below.

-   [Observation Well
    Attributes](http://pub.data.gov.bc.ca/datasets/179325/GW_Well_Attributes.csv)
    (License: [B.C.
    OGL](http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf)).
    [Dataset
    Details](http://www.data.gov.bc.ca/dbc/catalogue/detail.page?config=dbc&P110=recorduid:179325)

-   [Observation Well Groundwater Level
    Data](http://pub.data.gov.bc.ca/datasets/179324/GWL_monthly.csv)
    (License: [B.C.
    OGL](http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf)).
    [Dataset
    Details](http://www.data.gov.bc.ca/dbc/catalogue/detail.page?config=dbc&P110=recorduid:179324)

Published and Available On-Line at Environmental Reporting BC (`r updateDate`): <http://www.env.gov.bc.ca/soe/indicators/water/wells/index.html>

Appendix: Individual well maps and graphs
=========================================

Three plots were created for each well, including a map of the well’s
location (upper left).

The graph to the right of the map shows the monthly groundwater levels
relative to the annual average. This illustrates the seasonal nature of
water levels recorded in that well; many wells will have higher than
average water levels in the spring, and lower than average levels in the
late summer and fall. The shaded blue area shows the range of variation
within which 90% of observations in that month fall.

The bottom plot (called a "hydrograph") on each page shows the monthly
groundwater levels for the history of the well, with the orange line
representing the calculated trend in annual average groundwater levels.
The slope and significance of the trend are given below the title. Red
dots show missing values which were interpolated (see methods).

\newpage

```{r well_plots, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, results='asis', fig.height=7.5}
knitr_latex_char <- function(x) {
  y <- gsub("\\\\", "\\\\textbackslash{}", x) # backslash has to be first!
  y <- gsub("([#$%&_])", "\\\\\\1", y) # Doesdn't deal with { or } because of function{}
  y <- gsub("\\^", "\\\\textasciicircum{}", y)
  y <- gsub("~", "\\\\textasciitilde{}", y)
  return(y)
}

theme_set(base_theme)

for (reg in sort(unique(attr_wells$REGION_NM))) {
  cat(paste0('\\subsection{',reg,'}'))
  regdata <- attr_wells[attr_wells$REGION_NM==reg,]
  
  for (well in regdata$Well_Num) {
    if (is.na(well)) next
    
    well.attr <- regdata[regdata$Well_Num == well,]
    wellLong <- well.attr$Long
    wellLat <- well.attr$Lat
    wellColour <- with(well.attr, ifelse(category == "Large rate of decline"
                                         , colour.scale[3]
                                         , ifelse(category=="Stable or Increasing"
                                                  , colour.scale[1]
                                                  , colour.scale[2])))
    wellname <- well.attr$Well_Name
    
        
    cat(paste0('\\subsubsection*{', 
               knitr_latex_char(wellname),
               '}'))
    
    cat(paste0("This well is drilled into a ", well.attr$Aquifer_Type, 
               " aquifer, to a depth of ", well.attr$wellDepth_m, 
               "m and has a period of record from ", 
               format(well.attr$start_date, format="%B %d %Y"), 
               " to ", format(well.attr$last_date, format="%B %d %Y"),
               " \\newline "))
    
    plotdata <- monthly.data[monthly.data$Well_Num == well,]
    
    mapplot <- plotPointWithInset(long=wellLong, lat=wellLat
                                  , pointColour=wellColour
                                  , bigMap=wellMaps[well][[1]]
                                  , overviewMap=ggMapBC
                                  , overviewExtent=BCextent
                                  , opts=theme(plot.margin=unit(c(0.5,0.5,0.5,0.01)
                                                                ,"cm")))
    
    areaplot <- gwlAreaPlot(data=plotdata, trend=well.attr$trend_line_slope
                            , intercept=well.attr$trend_line_int
                            , state=well.attr$state, sig=well.attr$sig
                            , showInterpolated=TRUE, save=FALSE
                            , mkperiod="monthly"
                            , opts=theme(plot.margin=unit(c(0.5,0.5,0.5,0.01)
                                                          ,"cm")))
    
    monthplot <- gwlMonthlyPlot(dataframe=plotdata, splines=TRUE, save=FALSE
                                , opts=theme(plot.margin=unit(c(0.5,0.01,0.5,0.01)
                                                              ,"cm")))
    
    multiplot(mapplot + theme(text=element_text(size=10))
              , monthplot + theme(text=element_text(size=10)
                                  , axis.title.y=element_text(hjust=0.3))
              , areaplot + theme(text=element_text(size=10))
              , layout=matrix(c(1,2,3,3), nrow=2, byrow=TRUE)
              #, title=wellname
              )
    
    cat('\\newpage ')
  }

}
```
