---
topic: "water"
title: "Long-Term Trends in Groundwater Levels in B.C."
bibliography: gwl.bib
output: envreportutils.internal::print_ver
header-includes:
  - \usepackage{xcolor,colortbl}
  - \usepackage{longtable, booktabs}
---
<!--
Copyright 2018 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->


```{r setup, echo = FALSE, warning = FALSE, message = FALSE, cache = FALSE}

library(dplyr)
library(ggplot2)
library(grid)
library(RColorBrewer)
library(scales)
library(bcgroundwater)
library(envreportutils)
library(gridExtra)
library(xtable)
library(lubridate)
library(forcats)

knitr::opts_chunk$set(cache = FALSE)

base_theme <- theme_get()

# load('../tmp/clean_well_data.RData')
# load('../tmp/analysis_data.RData')
load('../tmp/results_viz.RData')
#load('../tmp/well_plots.RData')
load('../tmp/figures.RData')

attr_wells <- results_viz

state.summary <- table(attr_wells$state)
state.summary <- cbind(n = state.summary, prop = prop.table(state.summary))

theme_set(theme_classic() +
            theme(text = element_text(colour = "black"),
                  axis.line = element_blank(),
                  axis.ticks = element_blank(),
                  panel.grid.major = element_line(colour = "grey85", size = 0.5,
                                                linetype = 1),
                  panel.grid.minor = element_line(colour = "grey90", size = 0.5,
                                                linetype = 1),
                  panel.grid.major.x = element_blank(),
                  panel.spacing = unit(0.6, "lines"),
                  plot.title = element_text(vjust = 2, hjust = 0.5),
                  axis.title = element_text(vjust = 0.1),
                  legend.position = "bottom", legend.title = element_blank(),
                  legend.text = element_text(size = 12),
                  axis.text.x = element_blank(),
                  strip.background = element_blank()))

label.colour <- "black" 
colour.scale <- brewer.pal(3,"Blues")

updateDate <- format(Sys.Date(), "%B %Y")
```


Up to 1 million British Columbians are estimated to consume groundwater,
and hundreds of groundwater aquifers provide water for industries,
municipalities, farms, and rural homeowners in British Columbia. B.C.
operates a provincial observation well network&mdash;established in 1961&mdash;to monitor groundwater availability in areas of
high human use.

-   Observation wells are not used for domestic or commercial use, but
instead provide monitoring information on groundwater levels over time.

-   Monitoring groundwater levels allows us to know how much groundwater
is available given human use patterns, aquifer characteristics,
weather and climate patterns^[1](#fn1)^.

-   This indicator presents an analysis of long-term trends
in groundwater levels recorded at `r nrow(attr_wells)` observation wells that have been monitored for ten years or more and that were active as of 2009.

-  Groundwater level trends are summarized using three long-term trend categories: Stable or Increasing, Moderate Rate of Decline (3 to 10 cm/year), and Large Rate of Decline (more than 10 cm/year).


\newpage

## Provincial Summary of Trends in Groundwater Levels

-   Of the `r nrow(attr_wells)` examined observation wells, `r round(sum(state.summary[c("Stable","Increasing"),"prop"]*100))`% have water levels that are stable or increasing (with `r state.summary["Increasing","n"]` wells showing increasing trends), `r round(state.summary["Moderate Rate of Decline","prop"]*100)`% of wells show a moderate rate of decline in water levels, and `r round(state.summary["Large Rate of Decline","prop"]*100)`% show a large rate of decline in water levels.


\vspace{60pt}


```{r bc-bar, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, fig.width = 5, fig.height = 6, fig.align = 'center'}

bc_bar_chart + theme(legend.position = "bottom",
                     legend.direction = "vertical")
```

\newpage

## Summary of Trends in Groundwater Levels by B.C. Natural Resource Regions


\vspace{60pt}


```{r region, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, fig.width = 6, fig.height = 7, fig.align = 'center'}

plot(regional_bar_chart)
```



\newpage

## Trends in Groundwater Levels Observed at Observation Wells

\vspace{30pt}


```{r overview_map, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, results = 'asis', fig.width = 6, fig.height = 6, fig.align = 'center'}

plot(summary_map)
```

\newpage


\definecolor{stable}{HTML}{`r substr(colour.scale[1], 2, 7)`}
\definecolor{moderate}{HTML}{`r substr(colour.scale[2], 2, 7)`}
\definecolor{large}{HTML}{`r substr(colour.scale[3], 2, 7)`}

\small
```{r table, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, results = 'asis'}

out.table <- attr_wells[,c("REGION_NAME", "Well_Num", "start_date", "last_date",  
                            "trend_line_slope", "sig", "state")]

out.table$date_range <- paste(year(out.table$start_date), "-", 
                              year(out.table$last_date))

out.table <- arrange(out.table, REGION_NAME, as.numeric(Well_Num))

out.table$trend_line_slope <- -out.table$trend_line_slope

out.table <- out.table %>%
  select(REGION_NAME, Well_Num, date_range, sig, trend_line_slope,  state)

names(out.table) <- c("Region", "Well", "Date Range", 
                      "Sig", "Slope (m/yr)", "Category")

out.table$Sig <- ifelse(out.table$Sig < 0.05, "p < 0.05", 
                        round(out.table$Sig,2))

out.table$Well <- round(as.numeric(out.table$Well), 0)

x.table <- xtable(out.table, digits = c(0,0,0,0,3,3,0))

# Set the slope column as fixed width
align(x.table) <- c('l', rep('l', 4), 'p{0.5in}', rep('l', 1))

# Define colours of rows based on state
HL_col <- ifelse(out.table$Category == "Large Rate of Decline", "\\rowcolor{large}", 
                 ifelse(out.table$Category == "Moderate Rate of Decline", 
                        "\\rowcolor{moderate}", "\\rowcolor{stable}"))
HL_rows <- seq(0, length(HL_col)-1, by = 1)
HL_col[1] <- paste0("\\hline ", HL_col[1]) # Hack to get a hline below headings

print(x.table, tabular.environment = 'longtable', floating = FALSE,
      hline.after = c(-1,nrow(x.table)), 
      add.to.row = list(pos = list(as.list(HL_rows))[[1]], 
                        command = HL_col), 
      booktabs = TRUE, 
      include.rownames = FALSE, sanitize.colnames.function = NULL, 
      sanitize.text.function = NULL, comment = FALSE)
```


## More About Groundwater Levels

Groundwater levels are sensitive to precipitation, aquifer storage capacity, recharge rate&mdash;the rate at which surface water trickles down to refill a groundwater aquifer&mdash;and human withdrawal. Groundwater level trends presented here indicate long-term changes in water level, but have not been corrected for changes in precipitation patterns or other factors. Thus, any significant trends are not necessarily directly attributable to human use. However, information on long-term trends can be useful for prompting further research and informing decision-making. 


## References and Other Useful Links

-   Visit the [Monitoring Aquifers: Provincial Observation Well Network homepage](https://www2.gov.bc.ca/gov/content?id=B03D0994BB5C4F98B6F7D4FD8610C836) for more information on groundwater monitoring observation wells.

-   [Groundwater Wells and Aquifers (GWELLS)](https://apps.nrs.gov.bc.ca/gwells/)

-   [Water Data & Tools for British Columbia](https://www2.gov.bc.ca/gov/content?id=A9E4D7847FBC42CEB3A9E8ABAC364BD6)

-   [Previous B.C. State of Environment reports and indicators on groundwater levels in B.C.](https://www2.gov.bc.ca/gov/content?id=07DD1DCE7F784EFEBA5A26C32CB1646D)

- ^1^[Gurdak, J.J., R.T. Hanson, and T.R. Green. 2009. Effects of Climate Variability and Change on Groundwater Resources of the United States. United States Geological Survey. Fact Sheet 2009–3074. (PDF)](https://pubs.usgs.gov/fs/2009/3074/pdf/FS09-3074.pdf)


## Data

\*By accessing these datasets, you agree to the licence associated with
each file, as indicated in parentheses below.

-   [Indicator Summary Data: Long-term Trends in B.C. Groundwater Levels](https://catalogue.data.gov.bc.ca/dataset/a74f1b97-17f7-499b-84e7-6455e169e425)
(Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))

-   [Provincial Groundwater Observation Well Network - Groundwater Levels Data](https://catalogue.data.gov.bc.ca/dataset/57c55f10-cf8e-40bb-aae0-2eff311f16855)
(Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))


----

Published and Available On-Line at Environmental Reporting BC (`r updateDate`): <http://www.env.gov.bc.ca/soe/indicators/water/groundwater-levels.html>

Email correspondence to: envreportbc@gov.bc.ca

*Suggested Citation*:  
Environmental Reporting BC. 2019.  Long-term Trends in Groundwater Levels in B.C. State of Environment Reporting, Ministry of Environment and Climate Change Strategy, British Columbia, Canada.


\vspace{50pt}

## Appendix: Detailed Maps & Graphs for Groundwater Observation Wells Included in the Analysis

Three plots were created for each examined groundwater observation well, including a map of the well’s location (upper left).

The graph to the right of the map shows monthly groundwater levels
relative to the annual average water level. This illustrates the seasonal nature of
water levels recorded in that well&mdash;many wells will have higher than
average water levels in the spring, and lower than average levels in the
late summer and fall. The shaded blue area shows the range of variation
within which 90% of water level observations in that month fall.

The bottom plot&mdash;called a "hydrograph"&mdash;on each page shows the monthly
groundwater levels for the history of the well, with light grey
dots show missing values which were interpolated (see Methods). The trend category is provided below the title&mdash;if the trend is significant, based on a combination of statistical and environmental significance criteria&mdash;the calculated change in annual average groundwater levels is provided (meters per year) and visually represented by an orange trend line.


\newpage


```{r well_plots, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, results = 'asis', fig.height = 7.5, dev = "pdf"}

load('../tmp/well_plots.RData')

knitr_latex_char <- function(x) {
  y <- gsub("\\\\", "\\\\textbackslash{}", x) # backslash has to be first!
  y <- gsub("([#$%&_])", "\\\\\\1", y) # Doesn't deal with { or } because of function{}
  y <- gsub("\\^", "\\\\textasciicircum{}", y)
  y <- gsub("~", "\\\\textasciitilde{}", y)
  return(y)
}

theme_set(base_theme)

for (reg in sort(unique(attr_wells$REGION_NAME))) {
  cat(paste0('\\subsection{',reg," Region",'}'))
  regdata <- filter(attr_wells, REGION_NAME == reg)
  
  for (well in regdata$Well_Num) {
    if (is.na(well)) next
    
    wellname <- filter(regdata, Well_Num == well) %>%
      pull(Well_Name)
    
    cat(paste0('\\subsubsection*{',reg,": ",
               knitr_latex_char(wellname),
               '}'))
    
    # cat(paste0("This provincial groundwater observation well is drilled into a ",
    #            well.attr$AQUIFER_TYPE, 
    #            " aquifer to a depth of ", well.attr$wellDepth_m, 
    #            " metres. The well has a monitoring period of record from ", 
    #            format(well.attr$start_date, format = "%B %d %Y"), 
    #            " to ", format(well.attr$last_date, format = "%B %d %Y"), ". \\newline",
    #            "\\newline"))
    
    mapplot <- filter(well_plots, Well_Num == well) %>% pull(map_plot)
    monthplot <- filter(well_plots, Well_Num == well) %>% pull(month_plot)
    areaplot <- filter(well_plots, Well_Num == well) %>% pull(area_plot)
    
    grid.arrange(mapplot[[1]] + theme(text = element_text(size = 10),
                                 plot.margin = unit(c(0,0.5,0,0.01),"cm")),
                 monthplot[[1]] + theme(axis.text = element_text(size = 8),
                                   axis.title.x = element_blank(),
                                   plot.title = element_text(size = 10, face = "bold"),
                                   axis.title.y = element_text(size = 8,
                                                               hjust = 0.5,
                                                               vjust = 1,
                                                               face = "bold"),
                                   legend.box = "vertical",
                                   legend.position = "top",
                                   legend.text = element_text(size = 9),
                                   plot.margin = unit(c(1,0.01,1,0.01),"cm"),
                                   legend.spacing.y = unit(-.4, 'cm')),
                 areaplot[[1]] + theme(axis.text = element_text(size = 8),
                                  axis.title.x = element_blank(),
                                  axis.title.y = element_text(size = 8,
                                                              hjust = 0.5,
                                                              vjust = 1,
                                                              face = "bold"),
                                  legend.text = element_text(size = 9),
                                  legend.position = ("bottom"),
                                  plot.title = element_text(size = 11, face = "bold"),
                                  plot.subtitle = element_text(size = 9),
                                  plot.margin = unit(c(0.5,0.5,0.5,0.01),"cm")),
                 layout_matrix = matrix(c(1,2,3,3), nrow = 2, byrow = TRUE))
    
    cat('\\newpage ')
  }
  
}
```


## Methods

[**R**](http://www.r-project.org) package and code: We have developed an R package to facilitate working with, analyzing, and visualizing British Columbia groundwater level data. Download the 'bcgroundwater' package from [GitHub](https://github.com/bcgov/bcgroundwater). The source code for repeating the analysis presented on this page is also available on [GitHub](https://github.com/bcgov/groundwater-levels-indicator).


### Data

The data sets are all sourced from the B.C. Data Catalogue, distributed under the [Open Government Licence - British Columbia](https://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61).

-  Provincial Groundwater Observation Well Network - Groundwater Levels Data: 
<https://catalogue.data.gov.bc.ca/dataset/57c55f10-cf8e-40bb-aae0-2eff311f1685>
- Groundwater well metadata, such as location, depth, and aquifer type were obtained
from the Ground Water Wells (Spatial View) Dataset:
<https://catalogue.data.gov.bc.ca/dataset/e4731a85-ffca-4112-8caf-cb0a96905778>
- Natural Resource (NR) Regions, accessed with the ['bcmaps' R package](https://cran.r-project.org/web/packages/bcmaps/index.html): <https://catalogue.data.gov.bc.ca/dataset/dfc492c0-69c5-4c20-a6de-2c9bc999301f>


### Trend Analysis

Trend analysis was conducted using monthly groundwater levels from each
groundwater observation well that had ten or more years of data, fewer
than 25% of monthly observations missing, and was active as of January 1,
2008. This resulted in `r nrow(attr_wells)` wells with suitable data records for analysis. 

For each month that had more than one observation, the median was taken. After the time series were constructed, each was
inspected for consecutive missing values near the
beginning or the end of the time series, due to the strong influence of
these regions on trend analysis. If there were more than four
consecutive missing values in the first or last 10% of the time series,
the time series was truncated to remove the missing values. Remaining 
missing values were interpolated by first fitting a 12-period (i.e.,
monthly) basic structural time series model (BSM) to the data
[@harvey1990forecasting]. Fixed-interval smoothing
[@DurbinKoopman2001ts] was then applied to the fitted model to generate
predicted values for the entire time series; the predicted values were
substituted into the orignal time series where there were missing data.
This resulted in a full monthly time series for each well, with one
observation per month.  Finally, the
mean annual groundwater level was calculated for each well, and trend
analysis was performed using the annual values.

Each groundwater-level time series was analyzed for trend using
Trend-Free Pre-Whitening to remove lag-1 autocorrelation, which can
artificially increase the probability of detecting a significant trend
[@YuePilon2002autocorr]. In this method, the slope of each time series
is estimated using the Theil-Sen approach [@sen1968slope]. If the
estimated slope is different from zero, then the data are detrended by
the slope and the AR(1) 1s calculated for the detrended time series. The
residuals and the trend are combined and then tested for significance
using the Mann-Kendall trend test. These methods were implemented using
the R package ‘zyp’ [@pkg_zyp].

## Observed Long-term Trend Categories

The observed long-term trends were categorized using a combination of statistical and environmental significance criteria. Based on both the significance and magnitude of the calculated trend, trends in water levels in wells were classified as exhibiting the following categories:

-   **Stable or Increasing:** the trend was not significantly different from zero or the trend was significantly different from zero with an increasing trend or the trend was significantly different from zero with a declining trend of less than 3 cm per year.
-  **Moderate Rate of Decline:** the trend was significantly different from zero with a declining trend of between 3 and 10 cm per year.
-  **Large Rate of Decline:** the slope was significantly different from zero with a declining trend of more than 10 cm per year.

Time series were not corrected for precipitation or any other climatic
or hydrologic processes. As such, no causation for trends are implied.

### References
