---
topic: "water"
title: "Long-Term Trends in Groundwater Levels in B.C."
output: envreportutils.internal::print_ver
header-includes:
  - \usepackage{xcolor,colortbl}
  - \usepackage{longtable, booktabs}
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<!--
Copyright 2018 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->
```
```{r setup, echo = FALSE, warning = FALSE, message = FALSE, cache = FALSE}

library(dplyr)
library(ggplot2)
library(grid)
library(RColorBrewer)
library(scales)
library(bcgroundwater)
library(envreportutils)
library(gridExtra)
library(xtable)
library(lubridate)
library(forcats)
library(here)
library(magick)
library(png)
library(cowplot)

knitr::opts_chunk$set(cache = FALSE)

# load('../tmp/clean_well_data.RData')
# load('../tmp/analysis_data.RData')
load(here("tmp", "results_viz.Rdata"))
#load('../tmp/well_plots.RData')
load(here("tmp","figures.RData"))
load(here("tmp","input_regional.RData"))
load(here("tmp","by_aquifer.RData"))

leaflet_map = image_read(here("tmp","static_leaflet.png"))

attr_wells <- results_viz

state.summary <- table(attr_wells$state)
state.summary <- cbind(n = state.summary, prop = prop.table(state.summary))

theme_set(theme_classic() +
            theme(text = element_text(colour = "black"),
                  axis.line = element_blank(),
                  axis.ticks = element_blank(),
                  panel.grid.major = element_line(colour = "grey85", size = 0.5,
                                                linetype = 1),
                  panel.grid.minor = element_line(colour = "grey90", size = 0.5,
                                                linetype = 1),
                  panel.grid.major.x = element_blank(),
                  panel.spacing = unit(0.6, "lines"),
                  plot.title = element_text(vjust = 2, hjust = 0.5),
                  axis.title = element_text(vjust = 0.1),
                  legend.position = "bottom", legend.title = element_blank(),
                  legend.text = element_text(size = 12),
                  axis.text.x = element_blank(),
                  strip.background = element_blank()))

label.colour <- "black" 
colour.scale <- brewer.pal(3,"Blues")

updateDate <- format(Sys.Date(), "%B %Y")

```

Groundwater is an essential source of water for British Columbians. One
in four people in B.C. rely on groundwater for drinking water. It also
provides water for industry, municipalities, and farms.

Groundwater is found underground in layers of rock, gravel, or sand
called
[aquifers](https://www2.gov.bc.ca/gov/content/environment/air-land-water/water/groundwater-wells-aquifers/understanding-aquifers).
Human usage, weather, climate, and aquifer type all impact groundwater
levels. B.C. has a [network of \> 200 observation
wells](https://www2.gov.bc.ca/gov/content?id=B03D0994BB5C4F98B6F7D4FD8610C836)
to track groundwater availability in aquifers.These observation wells
are not used for domestic or commercial purposes.

This indicator shows a trend analysis of groundwater levels in
observation wells over time. The results are provided in a interactive
map below and in a static version of the indicator linked at the bottom
of the page. Summary figures of the trends observed across the province
and regionally are provided below the interactive map.

-   Wells that have been monitored for \> 10 years and were active as of
    2013 are analyzed in this indicator. There are `r nrow(attr_wells)`
    observation wells within the network that are assessed for changes
    over time. The location of the wells and the trend results are
    provided in the interactive map below.

-   The results are categorized by the statistical significance and
    magnitude of the change in groundwater levels over time. The
    categories are:

    -   **Increasing:** the trend was significantly different from zero
        with an increasing trend of 3 or greater cm per year.
    -   **Stable and/or Non-significant:** the trend was not
        significantly different from zero or the trend was significantly
        different from zero with a trend (increasing or decreasing) of
        less than 3 cm per year.
    -   **Moderate Rate of Decline:** the trend was significantly
        different from zero with a declining trend of between 3 and 10
        cm per year.
    -   **Large Rate of Decline:** the slope was significantly different
        from zero with a declining trend of more than 10 cm per year.
    -   **Insufficient Data:** there is an observation well but there is
        not enough data to calculate.

-   The location and status for each observation well can be seen in the
    map below. The observed long-term trends were categorized using a
    combination of statistical and environmental significance criteria.

-   Additional options to view trends are available in the interactive
    map in the online version of the indicator page. There are several
    different options to look at trends at observations wells

    -   Trend type: The user can select annual or monthly trends. The
        figures and map below are based on annual trends of groundwater
        levels.
    -   Time-span: The user can select the entire length of available
        data or can look at more recent trends over the past 10 or 20
        years. The figures and map below include the entire length of
        the dataset which varies from well to well. Please note that not
        all wells will have enough data for the 20-year time-span
        option. Also, by selecting a shorter time-span, the user is
        choosing to exclude part of the available dataset.

\newpage

\vspace{40pt}

# Trends in Groundwater Levels Observed by Aquifers

## Provincial Summary of Trends in Groundwater Levels (10 year trends)

- The `r nrow(results_viz_10)` observation wells assessed are located in`r nrow(by_aquifer)`aquifers. The results of 10 year trend analysis for observation wells were grouped by aquifer. The 10 year results are used to compare across aquifers as each well has a different time span of data. 
-   Of the `r nrow(by_aquifer)` observation wells examined,
    `r round(sum(input_summary[c("Stable and/or Non-significant","Increasing"),"prop"]*100))`%
    have water levels that are stable, non-significant trends, or
    increasing (with `r input_summary["Increasing","n"]` wells showing
    increasing trends).
    `r round(input_summary["Moderate Rate of Decline","prop"]*100)`% of
    wells show a moderate rate of decline in water levels, and
    `r round(input_summary["Large Rate of Decline","prop"]*100)`% show a
    large rate of decline. Note that only wells with enough data for
    trend analysis are included in the following figures.
- As each 

\vspace{10pt}

```{r bc-bar, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, fig.width = 6, fig.height = 6, fig.align = 'left'}
bc_bar_chart
```

\newpage

\vspace{40pt}

## Regional Summary of Trends in Groundwater Levels

-   The number of aquifers we assessed in each [Natural Resource
    Regions](https://www2.gov.bc.ca/gov/content/industry/forestry/managing-our-forest-resources/ministry-of-forests-lands-and-natural-resource-operations-region-district-contacts)
    varies. The highest number of aquifers assessed were in the West Coast Region. The
    lowest number of aquifers is the Skeena and Omineca regions.

\vspace{10pt}

```{r region, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, fig.width = 6, fig.height = 7, fig.align = 'left'}
regional_bar_chart
```


\newpage

## Location and Status of Observation Wells

\vspace{10pt}

```{r overview_map, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, fig.align = 'left', out.width = "100%"}
# summary map

plot(prov_map)
```

\newpage

\vspace{40pt}

## More About Groundwater Levels

Groundwater levels are sensitive to rainfall, aquifer storage capacity,
recharge rate—the rate at which surface water trickles down to refill a
groundwater aquifer—and human use. Trends presented here show long-term
changes in groundwater, but have not been corrected for the above
factors. Thus, any significant trends are not necessarily due to human
use alone. However, information on long-term trends can be useful for
prompting further research and informing decision-making.

<!-- Groundwater levels are sensitive to precipitation, aquifer storage capacity, recharge rate&mdash;the rate at which surface water trickles down to refill a groundwater aquifer&mdash;and human withdrawal. Groundwater level trends presented here indicate long-term changes in water level, but have not been corrected for changes in precipitation patterns or other factors. Thus, any significant trends are not necessarily directly attributable to human use. However, information on long-term trends can be useful for prompting further research and informing decision-making.  -->

## Methods

[**R**](http://www.r-project.org) package and code: We have developed an
R package to facilitate working with, analyzing, and visualizing British
Columbia groundwater level data. Download the 'bcgroundwater' package
from [GitHub](https://github.com/bcgov/bcgroundwater). The source code
for repeating the analysis presented on this page is also available on
[GitHub](https://github.com/bcgov/groundwater-levels-indicator).

<!-- ### Data -->

<!-- The data sets are all sourced from the B.C. Data Catalogue, distributed under the [Open Government Licence - British Columbia](https://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61). -->

<!-- -  Provincial Groundwater Observation Well Network - Groundwater Levels Data:  -->

<!-- <https://catalogue.data.gov.bc.ca/dataset/57c55f10-cf8e-40bb-aae0-2eff311f1685> -->

<!-- - Groundwater well metadata, such as location, depth, and aquifer type were obtained -->

<!-- from the Ground Water Wells (Spatial View) Dataset: -->

<!-- <https://catalogue.data.gov.bc.ca/dataset/e4731a85-ffca-4112-8caf-cb0a96905778> -->

<!-- - Natural Resource (NR) Regions, accessed with the ['bcmaps' R package](https://cran.r-project.org/web/packages/bcmaps/index.html): <https://catalogue.data.gov.bc.ca/dataset/dfc492c0-69c5-4c20-a6de-2c9bc999301f> -->

### Trend Analysis

Trend analysis was conducted using groundwater observation wells that
had ten or more years of data, fewer than 15% of monthly observations
missing, and was active as of January 1, 2013. This resulted in
`r nrow(attr_wells)` wells with suitable data records for analysis.

Trend analyses were calculated both annually and monthly. For each month
that had more than one observation, the median was taken. Annual mean
groundwater levels were then calculated for each year that contained
values for all twelve months. After the time series were constructed,
each was inspected for consecutive missing values near the beginning or
the end of the time series, due to the strong influence of these regions
on trend analysis. If there were more than four consecutive missing
values in the first or last 10% of the time series, the time series was
truncated to remove the missing values. 

Remaining missing values were interpolated by first fitting a 12-period (i.e., monthly) basic
structural time series model (BSM) to the data (Harvey, 1990).
Fixed-interval smoothing (Durbin & Koopman, 2012) was then applied to
the fitted model to generate predicted values for the entire time
series; the predicted values were substituted into the original time
series where there were missing data. This resulted in a full monthly
time series for each well, with one observation per month. Finally, the
mean annual groundwater level was calculated for each well, and trend
analysis was performed using the annual values.

Each groundwater-level time series was analyzed for trend using
the Zhang et al (2000) method for removing autocorrelation from time series data.  After correcting for autocorrelation, a Mann-Kendall test was conducted. In this method, the slope of each time series is estimated using the Theil-Sen approach (Sen, 1968). 

*Important Changes*

These methods were implemented using the R package ‘zyp’ version 0.11-1 (Bronaugh, D. & Schoeneberg, A., 2023). The analysis described above follows the Zhang *et al* (2000) method, however previous versions of this indicator used the Yue-Pilon method (Yue *et al* 2002). The methods were modified after the Yue Pilon method was found to have very high Type-1 error rates (Buerger 2017, Zhang and Zwiers 2004). Type 1 error rates occur when the null hypothesis is falsely rejected and analysis finds significant differences where none existed. 

Time series were not corrected for precipitation or any other climatic
or hydrologic processes. As such, no causation for trends are implied.


<!-- Trend analysis was conducted using monthly groundwater levels from each -->

<!-- groundwater observation well that had ten or more years of data, fewer -->

<!-- than 25% of monthly observations missing, and was active as of January 1, -->

<!-- 2008. This resulted in `r nrow(attr_wells)` wells with suitable data records for analysis.  -->

<!-- For each month that had more than one observation, the median was taken. After the time series were constructed, each was -->

<!-- inspected for consecutive missing values near the -->

<!-- beginning or the end of the time series, due to the strong influence of -->

<!-- these regions on trend analysis. If there were more than four -->

<!-- consecutive missing values in the first or last 10% of the time series, -->

<!-- the time series was truncated to remove the missing values. Remaining  -->

<!-- missing values were interpolated by first fitting a 12-period (i.e., -->

<!-- monthly) basic structural time series model (BSM) to the data -->

<!-- [@harvey1990forecasting]. Fixed-interval smoothing -->

<!-- [@DurbinKoopman2001ts] was then applied to the fitted model to generate -->

<!-- predicted values for the entire time series; the predicted values were -->

<!-- substituted into the orignal time series where there were missing data. -->

<!-- This resulted in a full monthly time series for each well, with one -->

<!-- observation per month.  Finally, the -->

<!-- mean annual groundwater level was calculated for each well, and trend -->

<!-- analysis was performed using the annual values. -->

<!-- Each groundwater-level time series was analyzed for trend using -->

<!-- Trend-Free Pre-Whitening to remove lag-1 autocorrelation, which can -->

<!-- artificially increase the probability of detecting a significant trend -->

<!-- [@YuePilon2002autocorr]. In this method, the slope of each time series -->

<!-- is estimated using the Theil-Sen approach [@sen1968slope]. If the -->

<!-- estimated slope is different from zero, then the data are detrended by -->

<!-- the slope and the AR(1) 1s calculated for the detrended time series. The -->

<!-- residuals and the trend are combined and then tested for significance -->

<!-- using the Mann-Kendall trend test. These methods were implemented using -->

<!-- the R package ‘zyp’ [@pkg_zyp]. -->


## Useful Links

-   Visit the [Monitoring Aquifers: Provincial Observation Well Network
    homepage](https://www2.gov.bc.ca/gov/content?id=B03D0994BB5C4F98B6F7D4FD8610C836)
    for more information on groundwater monitoring observation wells.

-   [Groundwater Wells and Aquifers
    (GWELLS)](https://apps.nrs.gov.bc.ca/gwells/)

-   [Water Data & Tools for British
    Columbia](https://www2.gov.bc.ca/gov/content?id=A9E4D7847FBC42CEB3A9E8ABAC364BD6)

-   [Previous B.C. State of Environment reports and indicators on
    groundwater levels in
    B.C.](https://www2.gov.bc.ca/gov/content?id=07DD1DCE7F784EFEBA5A26C32CB1646D)

## Data

The data sets are all sourced from the B.C. Data Catalogue, distributed
under the [Open Government Licence - British
Columbia](https://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61).

-   [Indicator Summary Data: Long-term Trends in B.C. Groundwater
    Levels](https://catalogue.data.gov.bc.ca/dataset/a74f1b97-17f7-499b-84e7-6455e169e425)

-   [Provincial Groundwater Observation Well Network - Groundwater
    Levels
    Data](https://catalogue.data.gov.bc.ca/dataset/57c55f10-cf8e-40bb-aae0-2eff311f1685)

-   [Groundwater well metadata, such as location, depth, and aquifer
    type were obtained from the Ground Water Wells (Spatial View)
    Dataset](https://catalogue.data.gov.bc.ca/dataset/e4731a85-ffca-4112-8caf-cb0a96905778)

-   [Natural Resource (NR)
    Regions](https://catalogue.data.gov.bc.ca/dataset/dfc492c0-69c5-4c20-a6de-2c9bc999301f),
    accessed with the ['bcmaps' R
    package](https://cran.r-project.org/web/packages/bcmaps/index.html)

### References

-   Bronaugh, D. & Shoeneberg, A. (2009). Package ‘zyp: Zhang + Yue-Pilon Trends Package’.
    CRAN Repository.
    
-   Buerger, G., 2017. On trend detection. Hydrological Processes 31, 4039-4042.

-   Durbin, J., & Koopman, S. J. (2012). Time series analysis by state
    space methods (Vol. 38). OUP Oxford.

-   [Gurdak, J.J., R.T. Hanson, and T.R. Green. 2009. Effects of Climate
    Variability and Change on Groundwater Resources of the United
    States. United States Geological Survey. Fact Sheet 2009–3074.
    (PDF)](https://pubs.usgs.gov/fs/2009/3074/pdf/FS09-3074.pdf)

-   Harvey, A. C. (1990). Forecasting, structural time series models and
    the Kalman filter.

-   Sen, P. K. (1968). Estimates of the regression coefficient based on
    Kendall's tau. Journal of the American statistical association,
    63(324), 1379-1389.

-   Yue, S., Pilon, P., Phinney, B., & Cavadias, G. (2002). The
    influence of autocorrelation on the ability to detect trend in
    hydrological series. Hydrological processes, 16(9), 1807-1829.
    
-   Zhang, X., Vincent, L. A., Hogg, W. D., & Niitsoo, A. (2000). Temperature and precipitation trends in Canada during the 20th century. Atmosphere-ocean, 38(3), 395-429.

-   Zhang, X., Zwiers, F.W., 2004. Comment on "Applicability of prewhitening to eliminate the influence of serial correlation on the Mann-Kendall test" by Sheng Yue and Chun Yuan Wang. Water
Resources Research 40.


------------------------------------------------------------------------

Published and Available Online at Environmental Reporting BC
(`r updateDate`):
<http://www.env.gov.bc.ca/soe/indicators/water/groundwater-levels.html>

Email correspondence to:
[envreportbc\@gov.bc.ca](mailto:envreportbc@gov.bc.ca){.email}

*Suggested Citation*:\
Environmental Reporting BC. 2024. Long-term Trends in Groundwater Levels
in B.C. State of Environment Reporting, Ministry of Environment and
Climate Change Strategy, British Columbia, Canada.

\newpage

## Appendix 1

### Table 1: Summary of *annual* results of trend analysis of groundwater levels

Note that only wells with enough data for trend analysis are included.

\small

```{r table, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, results = 'asis'}

# observation well table
results_viz %>%
  mutate(nYears = round(nYears)) %>%
  select(`Well Number` = Well_Num,
         `Result` = state,
         `Aquifer Type`= Aquifer_Type,
         `Region Name` = region_name,
         `Years Monitored` = nYears,
         `Intercept` = trend_line_int,
         `Slope` = trend_line_slope,
         `p-value` = sig) %>%
  arrange(`Region Name`) %>%
  knitr::kable(format = "latex",
               booktabs = TRUE,
               longtable = TRUE,
               linesep = "",
               align = "l",
               # col.names = linebreak(column_names, align = "l"),
               escape = FALSE) %>%
  kableExtra::kable_styling(
    position = "left",
    latex_options = c("striped", "repeat_header"),
    stripe_color = "gray!15"
  )
```

\newpage

## Appendix 2:

### Detailed Maps & Graphs for Groundwater Observation Wells Included in the Analysis

Three plots were created for each examined groundwater observation well,
including a map of the well’s location (upper left).

The graph to the right of the map shows monthly groundwater levels
relative to the annual average water level. This illustrates the
seasonal nature of water levels recorded in that well—many wells will
have higher than average water levels in the spring, and lower than
average levels in the late summer and fall. The shaded blue area shows
the range of variation within which 90% of water level observations in
that month fall.

The bottom plot—called a "hydrograph"—on each page shows the monthly
groundwater levels for the history of the well, with light grey dots
show missing values which were interpolated (see Methods). The trend
category is provided below the title—if the trend is significant, based
on a combination of statistical and environmental significance
criteria—the calculated change in annual average groundwater levels is
provided (meters per year) and visually represented by an orange trend
line.

\newpage

```{r well_plots, echo=FALSE, error=FALSE, fig.height=8, message=FALSE, warning=FALSE, results='asis', eval=FALSE}

# individual well plots
load('../groundwater-levels-indicator/tmp/well_plots.RData')

knitr_latex_char <- function(x) {
  y <- gsub("\\\\", "\\\\textbackslash{}", x) # backslash has to be first!
  y <- gsub("([#$%&_])", "\\\\\\1", y) # Doesn't deal with { or } because of function{}
  y <- gsub("\\^", "\\\\textasciicircum{}", y)
  y <- gsub("~", "\\\\textasciitilde{}", y)
  return(y)
}

# theme_set(base_theme)

for (reg in sort(unique(attr_wells$region_name))) {
  cat(paste0('\\subsection{',reg," Region",'}'))
  regdata <- filter(attr_wells, region_name == reg)
  
  for (well in regdata$Well_Num) {
    if (is.na(well)) next
    
    wellname <- filter(regdata, Well_Num == well) %>%
      pull(Well_Name)
    
    aquifer_id <- filter(regdata, Well_Num == well) %>%
      pull(aquifer_id)
    
    cat(paste0('\\subsubsection*{',reg,": ",
               knitr_latex_char(wellname),
               '}'))
    
    cat(paste0('\\subsubsection*{',"Associated aquifer number: ",
               knitr_latex_char(aquifer_id),
               '}'))
    
    # cat(paste0("This provincial groundwater observation well is drilled into a ",
    #            well.attr$AQUIFER_TYPE, 
    #            " aquifer to a depth of ", well.attr$wellDepth_m, 
    #            " metres. The well has a monitoring period of record from ", 
    #            format(well.attr$start_date, format = "%B %d %Y"), 
    #            " to ", format(well.attr$last_date, format = "%B %d %Y"), ". \\newline",
    #            "\\newline"))
    well = as.integer(well)
    #mapplot <- filter(well_plots, Well_Num == well) %>% pull(map_plot)
    # mapplot = readPNG(here(paste0("tmp/pngs/",well,".png")))
    #   mapplot = ggdraw() +
    #   draw_image(mapplot)
    monthplot <- filter(well_plots, Well_Num == well) %>% pull(month_plot)
    areaplot <- filter(well_plots, Well_Num == well) %>% pull(area_plot)
    
    grid.arrange(#mapplot + 
                   #theme(
                     # text = element_text(size = 10),
                                 #plot.margin = unit(c(0.5,0.5,0.5,0.5),"cm")),
                 monthplot[[1]] + theme(axis.text = element_text(size = 8),
                                   axis.title.x = element_blank(),
                                   plot.title = element_text(size = 10, face = "bold"),
                                   axis.title.y = element_text(size = 8,
                                                               hjust = 0.5,
                                                               vjust = 1,
                                                               face = "bold"),
                                   legend.box = "vertical",
                                   legend.position = "top",
                                   legend.text = element_text(size = 9),
                                   plot.margin = unit(c(0.5,0.5,0.5,0.01),"cm"),
                                   legend.spacing.y = unit(-.4, 'cm')),
                 areaplot[[1]] + theme(axis.text = element_text(size = 8),
                                  axis.title.x = element_blank(),
                                  axis.title.y = element_text(size = 8,
                                                              hjust = 0.5,
                                                              vjust = 1,
                                                              face = "bold"),
                                  legend.text = element_text(size = 9),
                                  legend.position = ("bottom"),
                                  plot.title = element_text(size = 11, face = "bold"),
                                  plot.subtitle = element_text(size = 9),
                                  plot.margin = unit(c(0.5,0.5,0.5,0.01),"cm")),
                 layout_matrix = matrix(c(1,2,3), nrow = 3, byrow = TRUE))
    
    cat('\\newpage ')
  }
  
}
```
